Eleven years of student replication projects provide evidence on the correlates of replicability in psychology

### 2023-04-26 paper/manuscript.Rmd
work on intro

introduction outline:

@@

Why? Many approaches, no definitive answers.
* Why are rates what they are. Both want to know to fix but also to watch out / calibrate caution. Can be approached from theoretical, interventional, and correlational dimensions.
* Theoretical -- we know people do bad practices and can model how this leads to lower replicability. Requires things we already have ideas about
* Interventional -- try things they think will make it better (like not doing bad practices), see what happens. Not everything is intervention prone, need to have strong belief first. Expensive!
* Correlational -- markets/etc say there is some predictive value to be had, but don't say if it's transparent and explainable. RP:P. 
   - Note that it's correlational, so can't do causality, but still useful for prediction, and as fodder for theory & intervention. (Interplay between them)

@@

<!--

@@

Questions:
- mention primary outcome in intro? 
- do we want to use the term p(cumulative)?

TODO fill in cites
flesh out theoretical approaches section

introduction outline:

Things don't replicate that well and it feels like a problem for cumulative science 
A cumulative science requires foundational empirical results that can be build upon in theories and future experiments. These load-bearing results need to be reliable; when they are not robust, attempts to build on them waste resources, trying to find explanations and effects where there may be none.

@@

* Theoretical -- we know people do bad practices and can model how this leads to lower replicability. Requires things we already have ideas about

@@

What's missing from the conversation and why our approach is good. 
* There's been a lot of discussion around replicabilty and it's role and yadda yadda (much overfocused on RP:P results because that's what's available). Growing sophistication and valid points, but we think it misses the mark about what practicing scientists are concerned about.

What does replicability mean? Prior approaches to replicability have focused on interpreting results in terms of a potentially problematic estimand: the probability of a finding in the literature being somehow truly replicable. Critics have pointed out that "true" replicability may not be possible to estimate outside of a specific sample [@vanbavel2016] or even time period [@ramscar].

@@

<!-- good below here -->

>>>>

<!--introduction outline:

@@

Why? Many approaches, no definitive answers.
* Why are rates what they are. Both want to know to fix but also to watch out / calibrate caution. Can be approached from theoretical, interventional, and correlational dimensions.
* Theoretical -- we know people do bad practices and can model how this leads to lower replicability. Requires things we already have ideas about
* Interventional -- try things they think will make it better (like not doing bad practices), see what happens. Not everything is intervention prone, need to have strong belief first. Expensive!
* Correlational -- markets/etc say there is some predictive value to be had, but don't say if it's transparent and explainable. RP:P.

@@

-->
<!--

@@

# Introduction

A cumulative science requires foundational empirical results that can be build upon in theories and future experiments. These load-bearing results need to be reliable; when they are not robust, attempts to build on them waste resources, trying to find explanations and effects where there may be none.

@@

@@

Many of the discussions and reanalyses have raised questions about what the right metrics for measuring replicability are and what standards are reasonable to expect. Some of this contention is based around notions that failed replications are referenda on the truth of the original finding.

Prior approaches to replicability have focused on interpreting results in terms of a potentially problematic estimand: the probability of a finding in the literature being somehow truly replicable. Critics have pointed out that "true" replicability may not be possible to estimate outside of a specific sample [@vanbavel2016] or even time period [@ramscar].

@@

### 2023-04-26 paper/manuscript.tex
work on intro

introduction outline:

Things don't replicate that well and it feels like a problem for cumulative science
* For a cumulative science, we want (many) of the load-bearing empirical results that form the bases of theories to hold up.
* People were acting as if publication of results meant they were true/reliable, trying to build on things; caused crises when some prominent results failed to replicate. Called into question reliability of literature as a whole -- lead to a few measurements.
* Large scale replications of chunks of the literature, but aren't many (list and explain)
* These rates feel lower than we want

Why? Many approaches, no definitive answers.
* Why are rates what they are. Both want to know to fix but also to watch out / calibrate caution. Can be approached from theoretical, interventional, and correlational dimensions.
* Theoretical -- we know people do bad practices and can model how this leads to lower replicability. Requires things we already have ideas about
* Interventional -- try things they think will make it better (like not doing bad practices), see what happens. Not everything is intervention prone, need to have strong belief first. Expensive!
* Correlational -- markets/etc say there is some predictive value to be had, but don't say if it's transparent and explainable. RP:P.
- Note that it's correlational, so can't do causality, but still useful for prediction, and as fodder for theory \& intervention. (Interplay between them)

What's missing from the conversation and why our approach is good.
* There's been a lot of discussion around replicabilty and it's role and yadda yadda (much overfocused on RP:P results because that's what's available). Growing sophistication and valid points, but we think it misses the mark about what practicing scientists are concerned about.
* Don't want to build on something and fail a bunch trying to get your variant to work and then realize it just doesn't work. Replicability (as we mean it here) A useful framing is ``ability to do cumulative science'' -- close to what people mean, may disagree with many things
* How this approach differs and what it gets us
* (built in feasibility under reasonable / small resource constraints)
* interest based sampling
* functional approach sidesteps issues of ``truth'' (may even be reliable in a different environment, but would be)

Teaser:
* We do x,y,z, and find that larger effect sizes and within subjects designs are the two strongest correlates of replicability in this sample.

Questions:
- mention primary outcome in intro?
- switch out fig 4 for supp fig 2?
- do we want to use the term p(cumulative)?

TODO fill in cites
flesh out theoretical approaches section

introduction outline:

Things don't replicate that well and it feels like a problem for cumulative science
* For a cumulative science, we want (many) of the load-bearing empirical results that form the bases of theories to hold up.
For a cumulative science, the empirical results that form the bases for theories and
* People were acting as if publication of results meant they were true/reliable, trying to build on things; caused crises when some prominent results failed to replicate. Called into question reliability of literature as a whole -- lead to a few measurements.
A major concern in psychology, as in other fields, is that the published literature may not replicate as well as had previously been assumed. This is illustrated both by a few prominent findings that turn out to not replicate at all {[}ex. terror management theory; Klein et al. (\protect\hyperlink{ref-klein2022}{2022}){]}, and by large-scale replication projects such as RP:P which found a replication rate around 40\% for a sample of the psychology literature (\protect\hyperlink{ref-openscienceconsortium2015}{Consortium 2015}). This low replication rates seems undesirable and seems to run counter to the idea of a cumulative science.

\begin{itemize}
\tightlist
\item
  Large scale replications of chunks of the literature, but aren't many (list and explain)
  We are aware of three large-scale replication efforts replicating experimental results in the existing psychology literature. RP:P sampled roughly 100 studies from articles published in three top psychology journals in 2008 and found a overall replication rate of around 40\% (\protect\hyperlink{ref-openscienceconsortium2015}{Consortium 2015}). The ManyLabs series of replications investigating heterogeneity has sampled target studies that are short and compare only two conditions, which is not representative of the psychology literature as a whole. Across Many Labs 1-3, 29 of 51 target effects (57\%) were found to replicate (\protect\hyperlink{ref-klein2014}{Klein et al. 2014}, \protect\hyperlink{ref-ebersole2016}{Ebersole et al. 2016}, \protect\hyperlink{ref-klein2018}{Klein et al. 2018}). Camerer et al. (\protect\hyperlink{ref-camerer2018}{2018}) found a roughly 60\% replication rate when they replicated all 21 behaviors studies published in Nature and Science from 2010-2015 that did not require special populations or special equipment. Those are all the large-scale replication projects of experimental psychology results available at this time. These roughly 170 results are the primary empirical results discussions of replication in pschology have to rely are.
\item
  These rates feel lower than we want
\end{itemize}

Why? Many approaches, no definitive answers.
What makes the replication rates in psychology so low? Knowing what predicts and drives non-replications would let us potentially increase the replication rate, and also calibrate our levels of caution and skepticism toward results. Theoretical, interventional and correlational approaches do not yet provide comprehensive answers.

\begin{itemize}
\tightlist
\item
  Theoretical -- we know people do bad practices and can model how this leads to lower replicability. Requires things we already have ideas about
\end{itemize}

Theoretical approaches modelling how questionable research practices (which many psychologists admit to) and a bias towards positive results would lead to a high rate of (non-replicable) false positives in the literature (CITE). There's some disconnect between these theoretically causal factors and empirical results about what causes or correlates with replicability.

\begin{itemize}
\item
  Interventional -- try things they think will make it better (like not doing bad practices), see what happens. Not everything is intervention prone, need to have strong belief first. Expensive!
  A couple studies have attempted to intervene on the low replication rate, but seeing if certain factors can affect whether studies replicate. For instance, across 16 studies, Protzko et al. (\protect\hyperlink{ref-protzko2020}{2020}) showed that better methodological practices, such as transparency, large sample sizes, and confirmatory tests, led to replication rates that matched theoretical expectations based on the effect sizes and sample sizes, and replication effect sizes comparable with the original. Many Labs 5 examined how adding expert advice to a replication process might intervene to improve the replicate rate, and found that it mostly didn't, at least on the 10 studies they looked at (\protect\hyperlink{ref-ebersole2020}{Ebersole et al. 2020}). These types of experiments are valuable, but they aren't very scalable because they are expensive and some potential influences on non-replication are not experimentally manipulable.
\item
  Correlational -- markets/etc say there is some predictive value to be had, but don't say if it's transparent and explainable. RP:P.
  Much of the work looking at predictors of replicability has been correlational. Prediction markets and elicitations have established that people can predict what studies will replicate above chance (\protect\hyperlink{ref-dreber2015}{Dreber et al. 2015}, \protect\hyperlink{ref-camerer2018}{Camerer et al. 2018}, \protect\hyperlink{ref-forsell2019}{Forsell et al. 2019}, \protect\hyperlink{ref-hoogeveen2019}{Hoogeveen et al. 2019}), but have not identified many concrete predictors that differentiate replications from non-replications. Machine learning approaches CITE similarly show predictive signal, but do not provide explainable predictor variables. In the RP:P sample, studies in cognitive psychology (as opposed to social psychology) and studies with larger effect sizes and smaller p-values were more likely to replicate CITE . All of these correlational approaches depend on data from replications, generally drawing heavily from the same small set of large-scale replication data points. In particular, the RP:P dataset itself is much discussed and reanalyzed (\protect\hyperlink{ref-anderson2016}{Anderson et al. 2016}, \protect\hyperlink{ref-etz2016}{Etz \& Vandekerckhove 2016}, \protect\hyperlink{ref-gilbert2016}{Gilbert et al. 2016}, \protect\hyperlink{ref-patil2016}{Patil et al. 2016}), so much of what we think we know about replicability may be overfit to the 100 studies included in in RP:P.

  \begin{itemize}
  \tightlist
  \item
    Note that it's correlational, so can't do causality, but still useful for prediction, and as fodder for theory \& intervention. (Interplay between them)
  \end{itemize}
\end{itemize}

What's missing from the conversation and why our approach is good.
* There's been a lot of discussion around replicabilty and it's role and yadda yadda (much overfocused on RP:P results because that's what's available). Growing sophistication and valid points, but we think it misses the mark about what practicing scientists are concerned about.

What does replicability mean? Prior approaches to replicability have focused on interpreting results in terms of a potentially problematic estimand: the probability of a finding in the literature being somehow truly replicable. Critics have pointed out that ``true'' replicability may not be possible to estimate outside of a specific sample (\protect\hyperlink{ref-vanbavel2016}{Van Bavel et al. 2016}) or even time period (\protect\hyperlink{ref-ramscar}{Ramscar et al. n.d.}).

@@

\begin{figure}[ht]
\includegraphics[width=1\linewidth,height=0.25\textheight]{manuscript_files/figure-latex/mod-results-1} \caption{Coefficient estimates and uncertainty from a model predicting subjective replication scores from the full dataset.}\label{fig:mod-results}
\end{figure}

>>>>

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

A cumulative science requires foundational empirical results that can be build upon in theories and future experiments. These load-bearing results need to be reliable; when they are not robust, attempts to build on them waste resources, trying to find explanations and effects where there may be none.

Scientists, policymakers, and the public often treat publication as a signal that a result is (likely to be) true. This trust in published findings leads scientists to attempt to build on them, but failures to build on them are explained away and rarely published, leaving the original results uncontested in the literature, while failures accumulate in private.

Publicized failures to replicate called into question the tacit assumption that most of the literature was robust and would replicate. Certain prominent findings turned out to not replicate at all {[}ex. terror management theory; Klein et al. (\protect\hyperlink{ref-klein2022}{2022}){]}, and a large scale replication project pegged the replication rate for findings in top-tier psychology journals at around 40\% (\protect\hyperlink{ref-openscienceconsortium2015}{Consortium 2015}).

Few large scale replications have been conducted in psychology, due to the resources required to run large numbers of replication studies. We are aware of three. RP:P sampled roughly 100 studies from articles published in three top psychology journals in 2008 and found a overall replication rate of around 40\% (\protect\hyperlink{ref-openscienceconsortium2015}{Consortium 2015}). ManyLabs investigated heterogeneity using short target studies that compared between two conditions each, but these studies are not representative of the psychology literature as a whole. Across Many Labs 1-3, 29 of 51 target effects (57\%) replicated (\protect\hyperlink{ref-klein2014}{Klein et al. 2014}, \protect\hyperlink{ref-ebersole2016}{Ebersole et al. 2016}, \protect\hyperlink{ref-klein2018}{Klein et al. 2018}). Camerer et al. (\protect\hyperlink{ref-camerer2018}{2018}) replicated all 21 behavioral social science studies from Nature and Science from 2010-2015 that were feasible; the replication rate was around 60\%. These roughly 170 replications of psychology studies are the primary empirical results on which discussions of replication rates in psychology are based.

A replication rate around 50\% runs counter to the prior assumption that most findings in the literature was robust enough to support cumulative science, and raises the question: What makes the replication rates in psychology so low?

Theoretical approaches to understanding the replication rate have modeled how flexible research practices, selective reporting, and a bias toward positive results could lead to a high rate of (non-replicable) false positives (CITE). However, while scientists admit to engaging in these practices, we don't yet know what proportion of non-replications are driven by these factors or how to identify these studies in the literature.

Experimental approaches attempt to intervene on the low replication rate, by seeing if changing certain factors affects whether studies replicate. Protzko et al. (\protect\hyperlink{ref-protzko2020}{2020}) showed, across 16 studies, that better methodological practices, such as transparency, large sample sizes, and confirmatory tests, led to replication rates that matched theoretical expectations based on the effect sizes and sample sizes, with replication effect sizes comparable with the original. Many Labs 5 examined how adding expert advice to a replication process might intervene to improve the replicate rate, and found that it mostly didn't, at least on the 10 studies they looked at (\protect\hyperlink{ref-ebersole2020}{Ebersole et al. 2020}). These types of experiments are valuable for testing potential causes of non-replication, but they don't scale well do to expense. Additionally, not all potential influences on non-replication are experimentally manipulable.

Correlational approaches looking for predictors of replicability are the most popular. Prediction markets and elicitations have established that people can predict what studies will replicate above chance (\protect\hyperlink{ref-dreber2015}{Dreber et al. 2015}, \protect\hyperlink{ref-camerer2018}{Camerer et al. 2018}, \protect\hyperlink{ref-forsell2019}{Forsell et al. 2019}, \protect\hyperlink{ref-hoogeveen2019}{Hoogeveen et al. 2019}), but have not identified many concrete predictors that differentiate replications from non-replications. Machine learning approaches CITE similarly show predictive signal, but do not provide explainable predictor variables. (\protect\hyperlink{ref-opensicenceconsortium2015}{\textbf{opensicenceconsortium2015?}}) looked for simple correlates of replicability in the RP:P sample and found that studies in cognitive psychology (as opposed to social psychology) and studies with larger effect sizes and smaller p-values were more likely to replicate. Correlations approaches depend on data from replications, generally drawing heavily from the same small set of data points. In particular, the RP:P dataset itself is much discussed and reanalyzed (\protect\hyperlink{ref-anderson2016}{Anderson et al. 2016}, \protect\hyperlink{ref-etz2016}{Etz \& Vandekerckhove 2016}, \protect\hyperlink{ref-gilbert2016}{Gilbert et al. 2016}, \protect\hyperlink{ref-patil2016}{Patil et al. 2016}) to the point that much of what we think we know about replicability may be overfit to the 100 studies included in in RP:P.

Many of the discussions and reanalyses have raised questions about what the right metrics for measuring replicability are and what standards are reasonable to expect. Some of this contention is based around notions that failed replications are referenda on the truth of the original finding.

Prior approaches to replicability have focused on interpreting results in terms of a potentially problematic estimand: the probability of a finding in the literature being somehow truly replicable. Critics have pointed out that ``true'' replicability may not be possible to estimate outside of a specific sample (\protect\hyperlink{ref-vanbavel2016}{Van Bavel et al. 2016}) or even time period (\protect\hyperlink{ref-ramscar}{Ramscar et al. n.d.}).

@@

\begin{figure}[ht]
\includegraphics[width=0.9\linewidth]{manuscript_files/figure-latex/mod-results-1} \caption{Coefficient estimates and uncertainty from a model predicting subjective replication scores from the full dataset.}\label{fig:mod-results}
\end{figure}
